# Aside: Understanding Methods and Generics in R {- .aside-chapter}

## You’ve Come a Long Way in R

By this point, you’ve learned a lot of R. You can wrangle data, fit models, make plots, write functions, etc. But you might also be starting to wonder: What’s R actually doing under the hood?

For example, how does `summary()` know what to do when you give it the output from `lm`? Or why does `plot()` sometimes give you a histogram and other times a scatterplot?

A good way to start answering those questions is to learn about **methods** and **generic functions**.

## Why Use Generics?

It’s nice when a function does the right thing depending on the kind of object you give it. You don’t want to memorize different function names for every kind of model. You just want to call `summary()`, `plot()`, or `print()` and get something useful. That’s what generic functions are for.

In R, many common functions like `summary()`, `plot()`, and `print()` are *generic* meaning they automatically call the version that matches the class of your object.

Let's run a linear model and use that as an example.

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
```

Everytime I run a linear model, I ask for a summary of it right away.

```{r}
summary(fit)
```

Note all the stuff in there! Somehow R calculated that from the object called `fit`. Let's look at that object. `fit` is a `list` with a bunch of stuff in it that authors of the `lm` function think will be useful to you. Here is its structure.

```{r}
str(fit)
```

That list has 12 things in it including the coefficients, residuals, fitted values and a lot more. And you can reach right in and grab them. A fitted model is just a list, and you pull pieces out by name with the `$` operator, the same way you would with any list or data frame.

```{r}
fit$coefficients
```

So the coefficients were sitting there all along. R also gives you a set of tidier accessor functions that do the same job:

```{r}
coefficients(fit)
```

Why bother with `coefficients()` when `fit$coefficients` already works? Because `coefficients()` is smart about what you hand it. It knows how to pull the coefficients out of an `lm`, but also out of dozens of other model types, and the author of each one decides what "the coefficients" should mean. Reaching in with `$` ties you to the exact internal layout of one kind of object. Asking with `coefficients()` works across all of them. (Oh, and R anticipated that users like me would be too lazy to type out `coefficients(fit)` so `coef(fit)` does the same thing.)

Not everything you'd want is sitting in that list, though. The t-statistics, p-values, and $R^2$ you'd report for a linear model aren't stored anywhere. `summary()` computes them on the fly from what is there. Let's revisit.

```{r}
summary(fit)
```

Ok. Check that out. We got a really nice formatted output that helps us interpret the linear model using the `summary` function. But note that the output of `summary(fit)` is really different than getting a summary of a numeric vector like the `mpg` in the `mtcars` data that went into the model.

```{r}
summary(mtcars$mpg)
```

### Same Function, Different Behavior. What Gives?

We used the same function (`summary`) and got two very different results.  

When we used `summary(fit)`, we're not calling some one-size-fits-all summary function. We're calling a specific **method** written for objects of class `"lm"`.

## How This Works

Every object in R carries a **class**, which is really just a label stuck to the object saying what kind of thing it is. Dispatch is R reading that label and acting on it. Here’s what R does when you call `summary(fit)`:

1. It checks the class of the object (`class(fit)`).
2. It looks for a function named `summary.classname`, in this case `summary.lm`.
3. If it finds that method, it runs it.
4. If not, it falls back to `summary.default()`.

This system is called S3. It's the oldest and most common object system in R, and it's what you'll encounter in almost every package you use. R has other systems too, S4 and R6 among them, and they dispatch a little differently. You'll bump into S4 with some spatial objects later in the book but won't even know that unless this kind of geekery interests you. For now, S3 covers the vast majority of what you'll see.

You can look under the hood:

```{r}
class(fit)
```

That tells you `fit` is of class `"lm"`. Now you can see all the methods that currently exist for summary in your environment. If you load more packages and you'll see more methods show up here.

```{r}
methods(summary)
```

And if you want to see the actual code that runs when you call `summary(fit)`, use `getS3method`:

```{r}
getS3method("summary", "lm")
```

Look that over. Try to find where the $R^2$ is calculated.

## What’s the Difference Between a Generic and a Method?

- A **generic function** is the function you call, like `summary()` or `plot()`.
  It doesn’t actually do the work itself. Instead, it decides which method to run based on the class of the object.

- A method is the specific function written for a particular class.  
  For example:
  - `summary.lm()` handles linear models.
  - `summary.data.frame()` handles data frames.
  - `summary.default()` is the fallback if no class-specific method is found.

You can look at the generic directly:

```{r}
summary
```

That UseMethod("summary") in the function body is what triggers dispatch. That is, it tells R to go find the right method.

## Common Generic Functions

Here are some other generic functions you're likely to run into:

- `print()` – display a human-readable version  
- `summary()` – provide a richer summary  
- `plot()` – show plots  
- `predict()` – make predictions  
- `residuals()` – return residuals  
- `fitted()` – return fitted values  
- `coef()` – extract coefficients  
- `update()` – refit a model with changes  
- `AIC()` / `BIC()` – compare model fit  
- `anova()` – compare nested models  
- `terms()` / `formula()` – extract parts of model structure  

## Exploring Methods

When you hit a function and want to know what's actually running, these are the tools, all of which you've seen in action above:

- `class(x)` to see what kind of object you're holding
- `methods(summary)` to see what classes a generic knows how to handle
- `methods(class = "lm")` to see what generics have a method for a class
- `getS3method("summary", "lm")` to read the actual code that runs

## Why This Matters

Understanding how generics and methods work helps you read other people's code, write better functions, and make sense of R's behavior when things feel a little magic.

You'll see this pattern come up throughout the book. When we work with spatial objects, things like `sf` features or `ppp` point patterns, the same idea applies. `plot()`, `summary()`, `print()` all do the right thing because someone wrote a method for that class.
