# Regression: GLS with Autocorrelated Residuals

## Big Idea
We flagged in the regression kriging module that spatially autocorrelated residuals from an OLS model are a problem. This is the module where we deal with that problem. It is also where we shift goals. Most of what we have done in this class has been interpolation: estimating values at unmeasured locations, with the prediction surface as the end product. Here we are doing something different. We want to understand the relationship between variables, estimate a coefficient that tells us how much $y$ changes with $x$, and test whether that relationship is distinguishable from noise. Any predictions are a byproduct; the coefficients and their uncertainty are the point.

Standard OLS regression assumes that the errors are independent. With spatial data, that assumption is commonly violated: nearby observations tend to share unmeasured spatial structure, which means their residuals are correlated rather than independent. When that happens, OLS coefficient estimates can be inefficient and their standard errors can be wrong, leading to overconfident inference. Generalized least squares (GLS) lets us explicitly model the spatial structure of the residuals and account for it, giving us better parameter estimates and valid hypothesis tests.

## Reading
@dormann2007methods provide a review of several methods for accounting for spatial structure in species distribution data. The paper does a good job of explaining the big picture on why space is important in a regression context. Much of the paper uses matrix notation that you might not be familiar with and the mathematical details are not critical. In this module, we will focus on using generalized least squares (GLS). For this paper, I suggest reading the intro and then skimming the rest. Pay attention to the first part of section three on spatial models based on GLS regression but don't sweat the details more than you want to.

@keitt2002accounting is also worth a heavy skim. They don't use GLS in their approach but do show how spatial methods outperform standard methods in a few case studies. Focus on the implications rather than the methods in that paper. I worked with Tim Keitt when I was a masters student and my goodness, he is a smarty pants. It's worth reading anything he's written.

## Packages
You'll want to install the packages below. We are going to fit a linear model using GLS as opposed to ordinary least squares and thus need `nlme`[@R-nlme] as well as some of our other spatial standards: `gstat`[@R-gstat], `spdep`[@R-spdep], `ncf`[@R-ncf], and `automap`[@R-automap]. As always, `tidyverse`[@R-tidyverse] for data wrangling.

```{r, message=FALSE}
library(automap)
library(gstat)
library(nlme)
library(spdep)
library(ncf)
library(tidyverse)
```

## Regression
A very common situation with environmental data is to want to do an ordinary least squares (OLS) regression using spatial data. If your data are explicitly spatial, you are doing spatial regression, and that has different pitfalls than standard regression. And the pitfalls matter here more than they did in the interpolation modules, because we care about getting the coefficients right. When we were kriging, a biased variance surface was a problem but the prediction surface might still have been useful. When we are trying to make inference about whether nitrogen drives tree growth, a coefficient with a wrong standard error and an inflated t-statistic is the whole problem.

The biggest issue is that if the residuals are autocorrelated, you have violated the assumption of [independence](https://en.wikipedia.org/wiki/Independent_and_identically_distributed_random_variables). When that happens, [the standard errors on the coefficient estimates tend to be underestimated, and the t-scores overestimated, when autocorrelation of the errors at short distances is positive](https://en.wikipedia.org/wiki/Autocorrelation#Regression_analysis). Any statistical inference could therefore be wrong: the p-values from a t-test will be artificially low. But despair not. If we take the spatial structure of the errors into account we can usually still model the association between $x$ and $y$ effectively.

Is this a little opaque? Let's make up an example. Say you want to build a model of tree growth as a function of plant-available nitrogen in the soil. Your idea is that trees with more nitrogen will grow better. You make a nice linear model $y=\beta_0 + \beta_1 x + \epsilon$ where $y$ is the basal area increment of the tree, $\beta_0$ is the intercept, $\beta_1$ is the coefficient on $x$ (plant-available N), and $\epsilon$ is the residual. You have a gorgeous dataset with 75 mapped trees, making this a spatial regression: $y_i=\beta_0 + \beta_1 x_i + \epsilon_i$ where $i$ is the location specified by northing and easting coordinates.

You run an OLS regression, get a positive slope, and a t-test says your model has skill (good $R^2$, $p<0.05$). Great. You fire off a letter to Nature, or write your thesis, or go dump some salmon carcasses on the trees, or just bathe in the satisfaction of a job well done.

But your model won't fit perfectly, so you have residuals ($\epsilon$). If those residuals are not independent and identically distributed (iid), you have violated the assumptions of OLS regression. The coefficients might be inefficiently estimated, the standard errors might be wrong, and the p-values might be inflated. Alas. The classic reason that residuals are not iid in spatial data is that they are not independent, which means autocorrelation.

In general, here is a good approach when dealing with spatial data in a regression.

1. Start by assuming iid (e.g., doing an ordinary least squares regression).

2. Look at the spatial structure of the residuals with Moran's I.

3. If the residuals look clean then carry on. Note that it's OK for $y$ or $x$ to be autocorrelated, what matters is the residuals. If the residuals have spatial structure, quantify it using Moran's I. Spatial structure might or might not be a problem. But if your life/thesis/soul depends on having a clean model you'll want to deal with it.

4. Fit a variogram to the residuals. We can use this information to fit a GLS model where we specify the spatial structure in the error term using a variogram model fit from the semivariance on the residuals.

## GLS and OLS
That four-step workflow makes GLS sound straightforward in outline. It helps to understand why GLS fixes what OLS gets wrong.

We have all been taught OLS. OLS is a **method** for estimating the unknown parameters in a linear regression model, e.g., the intercept ($\beta_0$) and slope ($\beta_1$) of $y = \beta_0 + \beta_1 x$. It works by finding the values that minimize the sum of squared errors. When the assumptions are met (errors are independent, equal variance, normal) the OLS estimators are optimal, often called BLUE: Best Linear Unbiased Estimators. OLS has a nice property that it can be taught without matrix algebra and connects naturally to ideas like correlation that most of us learn first in intro stats.

Here is the problem with autocorrelated errors. OLS treats every observation as carrying the same amount of independent information. But when errors are spatially autocorrelated, nearby observations share information: they are not fully independent. A positive residual at one location is likely to be accompanied by positive residuals nearby, not because of anything $x$ is doing, but because of unmeasured spatial structure in the system. OLS ignores this and effectively overcounts the information in the data. The practical consequence is that standard errors on the coefficient estimates tend to be underestimated when residuals are positively autocorrelated at short distances, which means t-statistics get inflated and p-values get too small. The coefficients themselves ($\hat\beta_0$, $\hat\beta_1$) may still be unbiased, but they are inefficiently estimated and any inference from the standard errors is suspect.

Generalized least squares (GLS) fixes this by explicitly accounting for the correlation structure of the errors. Instead of weighting each observation equally, GLS down-weights observations that are strongly correlated with their neighbors, because those observations are carrying less new information than they would if they were independent. In practice, this means we need to tell GLS what the correlation structure looks like, which is why we fit a variogram to the residuals first and pass that in. If the errors turn out to be independent, GLS simplifies and gives the same answer as OLS. OLS is a special case of GLS. The [OLS via Algebra and Matrices aside](07xAsideOLS-Algebra-Matrix.html) works through the matrix algebra if you want to see exactly how that works.

The practical upshot follows the [KISS principle](https://en.wikipedia.org/wiki/KISS_principle): use OLS when its assumptions are met, try GLS when they are not.

## Toy Example
Here is our goal: We will create a variable `y` that is a function of `x` but will have spatially autocorrelated noise ($\epsilon$). We will regress `y` on `x` and find to our horror that the residuals from this model are not iid. We will then apply a GLS model with an appropriate correlation structure and feel good about ourselves.

### Data
First, I'm going to make some autocorrelated noise $\epsilon$. This bit of code is not for the faint of heart. You can hum your way through it and know that at the end the variable $\epsilon$ will be spatially autocorrelated. If you want to follow along: we are making a variable $\epsilon$ that has spatial autocorrelation $\rho$ in the form $\epsilon = \rho W + e$ where $W$ is a weights matrix following an inverse distance weights process and $e$ is random noise ($N(0,1)$). We make `n` random points following complete spatial randomness (using `runif`) and then do some calculations to determine how each point is linked to others based on their distances (`dnearneigh` and `nbdists`). Based on the distances we make a weights matrix following IDW with power `p`. We get the spatial autocorrelation generating operator for $\rho$ via the weights matrix and then multiply by some white noise (`rnorm`) to get $\epsilon$. The details on this last step involve some matrix inversions that we don't need to unpack.

```{r,echo=FALSE}
set.seed(3515) #234 #104
```

```{r,message=FALSE, warning=FALSE}
# We will do n points
n <- 75

# CSR. Our mapped points.
easting <- runif(n,0,100)
northing <- runif(n,0,100)
points <- cbind(easting,northing)

# Generate the spatially autocorrelated error term, epsilon.
# Build neighbor objects with an inclusive distance threshold
dnb <- dnearneigh(x = points, d1 = 0, d2 = 150)
dsts <- nbdists(nb = dnb, coords = points)
# And follow an IDW process
p <- 2.25 # IDW power
idw <- lapply(dsts, function(x) x^-p)
# Weights matrix as a list
wList <- nb2listw(neighbours = dnb, glist=idw, style="W") 

# Compute SAR generating operator via a weights matrix
inv <- spatialreg::invIrW(x = wList, rho = 0.75)

# Calculate epsilon and rescale as z scores
epsilon <- inv %*% rnorm(n) 
epsilon <- scale(epsilon[,1])[,1]
```

OK, by great magic we now have a variable that is from random spatial points but has spatially autocorrelated values (that follow an IDW process). If you want, you can imagine that $\epsilon$ is a variable in nature that is spatially autocorrelated that we didn't measure (e.g., maybe soil moisture availability following the tree example above). How intense is the spatial autocorrelation?

```{r}
moran.test(epsilon, wList)
```

We are going to use $\epsilon$ as the error term in a model. I'll use the terms "errors" and "residuals" interchangeably throughout these notes.

We'll now make `x` and then $y=\beta_0 + \beta_1(x) + \epsilon$. We will run a linear model and look at the residuals of that model using Moran's I.

```{r}
# Generate x as noise.
x <- rnorm(n)
# Generate y as a function of x and epsilon 
# (remember epsilon is spatially autocorrelated but we don't know that)
B0 <- 3 # intercept
B1 <- 0.5 # slope
y <- B0 + B1*x + epsilon

# Make a data.frame
dat <- data.frame(easting,northing,y,x)
```

So in our thought experiment here are the data. The independent variable `x`:
```{r}
ggplot(data = dat, mapping = aes(x=easting,y=northing,
                                 size=x, fill=x)) + 
  geom_point(shape=21,alpha=0.7) +scale_fill_viridis_c() +
  guides(fill=guide_legend(), size = guide_legend()) +
  coord_fixed() +
  ggtitle("x")
```

And the dependent variable `y`:
```{r}
ggplot(data = dat, mapping = aes(x=easting,y=northing,
                                 size=y, fill=y)) +
  geom_point(shape=21,alpha=0.7) + scale_fill_viridis_c() + 
  guides(fill=guide_legend(), size = guide_legend()) + 
  coord_fixed() + ggtitle("y")
```

### Analysis
Following our imagined example above we have a dependent variable `y` (tree growth) that we want to model using an independent variable `x` (nitrogen). The points (trees) are mapped using coordinates (`easting` and `northing`) and we can imagine those are in meters. Because we made up these data we know that the residual term on this model is autocorrelated noise (`epsilon`). If you don't like having the errors be something theoretical in your head, you can imagine that this term is mostly some variable we haven't measured that might provide explanatory power (e.g., a topographic variable relating to soil moisture availability).

We will start by assuming `epsilon` follows an iid process (e.g., $N(0,1)$) and we will fit a model with uncorrelated errors. Here we use `gls` but this gives the same result as `lm` would because `gls` assumes iid errors unless told otherwise.

```{r}
# Fit a model using gls
glsNaive <- gls(y~x, dat)
summary(glsNaive)
```

We fit a linear model of the form $y=\beta_0 + \beta_1x$ using GLS and got estimates for the intercept ($\hat\beta_0$) and slope ($\hat\beta_1$). We assumed that the error term was uncorrelated. Looking at summary, we can see the estimate of the slope ($\hat\beta_1$) is `r round(summary(glsNaive)$tTable[2,1],3)` and significantly different from zero. The important thing to wrap your head around here is that we just fit GLS with uncorrelated errors, meaning it's essentially identical to OLS. Compare the above to the more familiar `summary(lm(y~x,data=dat))`.

Since these data are spatial, the key for us is to look at the residuals and see if they are autocorrelated. Let's map them first.

```{r}
# Add residuals to dat
dat$glsNaiveResids <- residuals(glsNaive,type="normalized")
ggplot(data = dat, mapping = aes(x=easting,y=northing,
                                 size=glsNaiveResids, fill=glsNaiveResids)) +
  geom_point(shape=21) +
  scale_fill_gradient2() +
  guides(fill=guide_legend(), size = guide_legend()) +
  coord_fixed()
```

From the map, the residuals look aggressively non-random. Positive values are near positive values and negative ones near negative values. Let's quantify that with a correlogram.

```{r}
# Test them for spatial autocorrelation using a correlogram.
residsI <- spline.correlog(x=dat$easting, y=dat$northing,
                           z=dat$glsNaiveResids, resamp=50, quiet=TRUE)
plot(residsI,xlim=c(0,max(dist(points))/3))
```

The residuals of this model are autocorrelated at short distances and we have violated the assumption of independent residuals. But do not despair. We will use a GLS model that allows us to specify a spatial correlation structure for the residuals and perform hypothesis testing in the presence of spatial structure. Doing so will give us new and improved estimates of the parameters. We will thus sleep well knowing in our hearts that we have done good.

We need to determine an appropriate theoretical model for the spatial structure of the residuals and pass that to `gls`. Before we do, it is worth being concrete about what happens under the hood when we do. GLS builds a variance-covariance matrix $\Sigma$ for the residuals, where each off-diagonal entry $\Sigma_{ij}$ is a function of the distance between locations $i$ and $j$ -- with the functional form coming from the correlation structure you specify. Then it estimates both the regression coefficients and the correlation parameters (the range, nugget, and partial sill) together by maximizing the likelihood. The estimator is $\hat\beta = (X^T \Sigma^{-1} X)^{-1} X^T \Sigma^{-1} y$, which you can compare to the OLS version $(X^T X)^{-1} X^T y$: OLS weights every observation equally, GLS weights by the inverse of the covariance structure. The default method in `gls` is REML, which gives better estimates of the variance parameters than plain ML, at the cost of not being able to compare models with different fixed effects via likelihood ratio tests. You generally want REML for estimation and ML if you need to compare fixed effects with `anova`.

In `nlme`, correlation structures are represented by a family of objects that all inherit from a parent class called `corStruct`. Think of `corStruct` as the generic container: it tells `gls` "here is the shape of the correlation among the residuals." The spatial flavor of this is `corSpatial`, which takes a `form` argument specifying your coordinate columns, a `nugget` flag, and a `type` argument for the model family. When you write `corSpatial(form=~easting+northing, nugget=TRUE, type="spherical")` you are building one of these objects. There are also convenience wrappers `corSpher`, `corGaus`, and `corExp` that are just shorthand for `corSpatial` with a fixed `type`. The reason the class hierarchy is worth knowing is that `gls` accepts any `corStruct` in its `correlation` argument, not just spatial ones -- `corAR1` for time series autocorrelation works the same way, for instance.

The families available in `gls` roughly match the theoretical variogram models we used with kriging: spherical, exponential, Gaussian, and so on. So, let's fit a variogram to the residuals and see which model fits best.

Here we'll cheat a little and use `autofitVariogram` to get a model suggestion rather than go through the fitting process by hand with `variogram`, `vgm`, and `fit.variogram` from `gstat`.

```{r}
datSF <- dat %>% st_as_sf(coords = c("easting","northing"))
# A variogram of the residuals
plot(autofitVariogram(glsNaiveResids~1, input_data = datSF, model = c("Gau","Sph","Exp")))
```

Look at which theoretical model `autofitVariogram` chose. Here is a frustrating mismatch worth knowing about. `gstat` can fit a lot of variogram models: spherical, exponential, Gaussian, Matern, Stein's Matern, power, and several others. `corSpatial` supports only five: `"spherical"`, `"exponential"`, `"gaussian"`, `"linear"`, and `"rational"`. No Matern. In practice, the linear model is rarely appropriate for spatial residuals, so you are really working with three sensible options. If `autofitVariogram` returns a Matern (model code `Ste` or `Mat`), you cannot plug that directly into `corSpatial`. Your options are to re-run `autofitVariogram` restricting the candidate models to `c("Gau","Sph","Exp")` -- which is what we did above -- or to look at the empirical variogram and make a judgment call about which of the three available families it most resembles. This is not ideal but it is the reality of working across two packages with different design philosophies.

We'll use the chosen model family as the `type` argument to `corSpatial` below. One important thing to understand: we are not passing the variogram parameter estimates (the range, nugget, sill) into `corSpatial`. The `type` argument just specifies the functional form of the correlation structure. The actual parameter values are estimated by `gls` itself during model fitting. The variogram step tells us what shape to use; `gls` figures out the scale.

```{r}
csSpatial <- corSpatial(form=~easting+northing,nugget=TRUE, type = "spherical")
glsUpdated <- update(glsNaive,correlation=csSpatial)
summary(glsUpdated)
```

We can then use `update` to re-fit the model with the new correlation structure. Note that this is the same as fitting a new model via `gls(model = y~x, data = dat, correlation = csSpatial)` but `update` is a cleaner way of making sure the model specification stays identical for the sake of comparison.

A word of warning before you get too comfortable. The optimizer that `gls` uses to estimate the correlation parameters can fail, or can produce results that look wrong. Common symptoms are a convergence warning, `NA` values in the summary output, or a log-likelihood that is worse than the naive model. This tends to happen when the correlation structure you specified doesn't fit the data well, when the nugget-to-sill ratio is close to zero or one (making the model nearly unidentifiable), or just when the likelihood surface is sufficiently flat that the optimizer wanders. If this happens, try a few things: switch to a different model family, try `nugget=FALSE` to simplify the model, or fit the model with `method="ML"` temporarily to check whether the REML surface is the culprit. It is also worth inspecting `Variogram(glsUpdated)` from `nlme` after a fit: this plots the theoretical correlation function from the fitted model against the empirical variogram of the normalized residuals, and is a good diagnostic for whether the optimizer landed somewhere sensible.

This is interesting. We will interpret more below but note that the slope coefficient is different from `glsNaive` to `glsUpdated`. Let's look at the spatial structure of the residuals from `glsUpdated`.

```{r}
# add the residuals to the spatial dat object
dat$glsResids <- residuals(glsUpdated,type="normalized")
# Map the residuals from gls
ggplot(data = dat, mapping = aes(x=easting,y=northing,
                                 size=glsResids, fill=glsResids)) +
  geom_point(shape=21) +
  scale_fill_gradient2() +
  guides(fill=guide_legend(), size = guide_legend()) +
  coord_fixed()
```

They don't look to be autocorrelated. But we should test them:

```{r}
residsI <- spline.correlog(x=dat$easting, y=dat$northing,
                           z=dat$glsResids, resamp=50, quiet=TRUE)
plot(residsI,xlim=c(0,max(dist(points))/3))
```

And the correlogram shows that they are clean -- autocorrelation has been accounted for.

### Interpretation
So what's the big deal? We started with a model that showed `x` as a significant predictor of `y`. We ended up, after much wailing and rending of garments, with a model that showed `x` as a significant predictor of `y`. Is it a better model? Was it worth the trouble? Let's consider a few things.

1. The GLS model with the correlated residuals has a lower AIC than the naive uncorrelated model (`r round(AIC(glsUpdated),0)` vs. `r round(AIC(glsNaive),0)`). AIC penalizes model complexity, so a lower AIC means the GLS model is explaining more of the variance in `y` without just adding parameters to do it. It is a  better fit, not just a more complicated one.

2. The original function we used to make $y$ was $y=\beta_0 + \beta_1x + \epsilon$ where $\beta_0$ = `r B0` and $\beta_1$ = `r B1`. Both GLS models got the intercept right (`r round(coef(glsUpdated)[1],2)` for `glsUpdated` and `r round(coef(glsNaive)[1],2)` for `glsNaive`). But the GLS model with the correlated residuals does a better job of recovering the true slope (`r round(coef(glsUpdated)[2],2)` vs `r round(coef(glsNaive)[2],2)`).

    You might notice that the naive slope (`r round(coef(glsNaive)[2],2)`) is further from the truth (0.5) than the GLS slope (`r round(coef(glsUpdated)[2],2)`). Earlier we said OLS is unbiased even with autocorrelated errors -- so why did the coefficient move? Unbiased means that if you ran this simulation thousands of times, the average OLS estimate would center on 0.5. GLS would also center on 0.5, but with less spread around it. That's what inefficiency means: not a systematic pull away from the truth, but more variability from sample to sample. In this particular random draw, OLS overshot and GLS happened to land closer. In a different draw it could go the other way. The advantage of GLS is not that it always wins in any single sample, but that its estimates are less variable across samples and, crucially, its standard errors are actually valid.

    The scenario where OLS can be truly biased (not just inefficient) is worth flagging: if $x$ is correlated with the spatial structure of $\epsilon$, meaning an important spatially-structured variable was omitted from the model. In that case the spatial autocorrelation in the residuals is partly confounded with the effect of $x$ and OLS will systematically misestimate the slope. In our simulation $x$ is purely random noise with no spatial structure, so this doesn't apply. But in real data, where your predictor often covaries with geography, it can.

3. The standard error on $\beta_1$ is lower in the model with correlated residuals than in the naive model (`r round(summary(glsUpdated)$tTable[2,2],4)` vs `r round(summary(glsNaive)$tTable[2,2],4)`). This might seem backwards: you might expect that accounting for autocorrelation would widen the uncertainty rather than narrow it. The direction of the change actually depends on the spatial configuration of the data and the pattern of autocorrelation. What's happening here is that by correctly partitioning the variance into a regression component and a spatial correlation component, GLS estimates $\beta_1$ more efficiently. The naive OLS model was attributing some of the spatially structured noise to the slope, which inflated the apparent uncertainty. GLS separates those two sources and gets a cleaner estimate of each.

Finally, we did it right, and that should be satisfying. What would have happened if $\beta_1$ had been smaller? That is, what if $y$ was made up of more noise and less $x$? E.g., $y=\beta_0 + 0.5(\beta_1)x + 2(\epsilon)$? This model, with `r n` points, should be pretty robust on the simple hypothesis test on the slope (H$_1$: $\beta_1 \neq 0$) but we know that parameter estimation is inefficient with autocorrelated residuals, so having clean residuals and good parameter estimation is important.

## Your work
The joy of simulated data is, of course, that we know the real answer going in. We knew that $y=\beta_0 + \beta_1x + \epsilon$ and we knew that the residual term was spatially autocorrelated. Obviously, that's a luxury we don't have when making inference about data we collect.

The file `birdDiv.csv` has data on bird diversity and plant community diversity (unitless, with higher numbers indicating greater diversity) and coordinates for 64, 1-ha plots. Coordinates are UTM and thus in meters. Your work is to model bird diversity as a function of plant diversity with the expectation that higher plant diversity is associated with higher bird diversity. Watch those residuals and interpret carefully!

```{r,eval=FALSE,echo=FALSE}

dat <- read.csv("birdDiv.csv")

glsNaive <- gls(birdDiv~plantDiv, dat)

summary(glsNaive)

dat$glsNaiveResids <- residuals(glsNaive,type="normalized")

residsIglsNaive <- spline.correlog(x=dat$UTME, y=dat$UTMN,xmax=8e04,
                            z=dat$glsNaiveResids, resamp=50, quiet=TRUE)
plot(residsIglsNaive)

# note that you can choose what models you fit in autofit vgm
datSF <- dat %>% st_as_sf(coords=c("UTME","UTMN"))
aVariogram <- autofitVariogram(glsNaiveResids~1,
                               input_data = datSF,
                               model = c("Gau","Exp","Sph"),
                               verbose = TRUE)
plot(aVariogram)
# what is the max distance to look at for autocorrelation?
# let's use 1/3 of the max point to point distance
max(dist(cbind(dat$UTME, y=dat$UTMN)))/3 # about 84km 

plot(spline.correlog(x=dat$UTME, y=dat$UTMN,xmax=84e03,
                     z=dat$glsNaiveResids, resamp=100, quiet=TRUE))
# def short distance ac

# try a bunch
csGaus <- corSpatial(form=~UTME+UTMN,nugget=TRUE,type="gaus")
csGausNoNugget <- corSpatial(form=~UTME+UTMN,nugget=FALSE,type="gaus")

glsGau <- update(glsNaive,correlation=csGaus)
glsGauNoNug <- update(glsNaive,correlation=csGausNoNugget)


csSph <- corSpatial(form=~UTME+UTMN,nugget=TRUE,type="sph")
csSphNoNug <- corSpatial(form=~UTME+UTMN,nugget=FALSE,type="sph")

glsSph <- update(glsNaive,correlation=csSph)
glsSphNoNug <- update(glsNaive,correlation=csSphNoNug)

csExp <- corSpatial(form=~UTME+UTMN,nugget=TRUE,type="exp")
csExpNoNug <- corSpatial(form=~UTME+UTMN,nugget=FALSE,type="exp")

glsExp <- update(glsNaive,correlation=csExp) # nope
glsExpNoNug <- update(glsNaive,correlation=csExpNoNug)

AIC(glsNaive,glsGau,glsGauNoNug,glsSphNoNug,glsExpNoNug)

# why no work? There is no sill!
anova(glsNaive,glsGau)
plot(Variogram(glsGau))
plot(Variogram(glsGauNoNug))
plot(Variogram(glsSph))
plot(Variogram(glsSphNoNug))
plot(Variogram(glsExp))
plot(Variogram(glsExpNoNug))


dat$glsSphResids <- residuals(glsSph,type="normalized")
dat$glsGauResids <- residuals(glsGau,type="normalized")


residsIglsSphResids <- spline.correlog(x=dat$UTME, y=dat$UTMN,xmax=8e04,
                            z=dat$glsSphResids, resamp=50, quiet=TRUE)
residsIglsGauResids <- spline.correlog(x=dat$UTME, y=dat$UTMN,xmax=8e04,
                            z=dat$glsGauResids, resamp=50, quiet=TRUE)

par(mfcol=c(1,2))
plot(residsIglsGauResids,main="GLS w/ corGaus Good!")
plot(residsIglsSphResids,main="GLS w/ corSper Bad!")

```
