# Solutions: Geostatistics III - Kriging

```{r, echo=FALSE, include=FALSE}
library(sf)
library(tidyverse)
library(gstat)
library(tmap)
```

```{r, message=FALSE}
precip_sf  <- readRDS("data/waPrecip.rds")
wa_sf      <- readRDS("data/waOutline.rds")
wa_grid_sf <- readRDS("data/waGrid.rds")
```

## 1. Empirical Variogram

```{r}
precip_var <- variogram(precip_mm ~ 1, data=precip_sf)
plot(precip_var)
```

The variogram rises steeply at short distances (high spatial autocorrelation nearby) and flattens around 200-300 km. There's a clear sill. The nugget is relatively small, suggesting most variation is spatially structured rather than noise.

## 2. Fit Three Models

```{r}
vExp <- fit.variogram(precip_var, vgm(psill=200000, model="Exp", range=150000, nugget=5000))
vSph <- fit.variogram(precip_var, vgm(psill=200000, model="Sph", range=200000, nugget=5000))
vGau <- fit.variogram(precip_var, vgm(psill=200000, model="Gau", range=100000, nugget=5000))

plot(precip_var, vExp, main="Exponential")
plot(precip_var, vSph, main="Spherical")
plot(precip_var, vGau, main="Gaussian")
```

Visually compare the fitted lines to the empirical points. The exponential model rises fastest at short distances and typically fits well when there's a smooth decay. Spherical levels off sharply at the range. Gaussian is the smoothest and can over-smooth near the origin.

## 3. Cross-Validate

```{r}
cv_krige <- function(model, data, grid) {
  krig <- krige.cv(precip_mm ~ 1, data, model=model)
  rmse <- sqrt(mean(krig$residual^2, na.rm=TRUE))
  r2   <- cor(krig$observed, krig$var1.pred, use="complete.obs")^2
  data.frame(RMSE=rmse, R2=r2)
}

bind_rows(
  cv_krige(vExp, precip_sf) %>% mutate(Model="Exponential"),
  cv_krige(vSph, precip_sf) %>% mutate(Model="Spherical"),
  cv_krige(vGau, precip_sf) %>% mutate(Model="Gaussian")
)
```

Exponential wins: R² = 0.902, RMSE = 138.2 mm. This is consistent with the typical behavior of precipitation fields -- the correlation decays more like an exponential than a Gaussian, and the spherical cutoff is too abrupt.

## 4. Kriging Prediction and Uncertainty

```{r}
krig_pred <- krige(precip_mm ~ 1, precip_sf, wa_grid_sf, model=vExp)

tm_shape(wa_sf) + tm_borders() +
tm_shape(krig_pred) +
  tm_raster(col="var1.pred", palette="Blues", title="OK (mm)") -> m1

tm_shape(wa_sf) + tm_borders() +
tm_shape(krig_pred) +
  tm_raster(col="var1.var", palette="YlOrRd", title="Kriging Variance") -> m2

tmap_arrange(m1, m2, nrow=1)
```

The prediction surface looks similar to IDW and TPS. The variance map is the real payoff: high kriging variance in data-sparse areas (northeast corner, high Cascades) and low variance where stations are dense. This is honest spatial uncertainty quantification. IDW gives you nothing like this.

## 5. Comparison

Ordinary kriging (R² = 0.902, RMSE = 138.2 mm) is almost identical to TPS (R² = 0.904, RMSE = 136.2 mm) in LOOCV skill. Both are substantially better than IDW (R² = 0.863, RMSE = 164.4 mm). The main practical advantage of kriging over TPS isn't prediction skill -- it's the kriging variance map. When you need to communicate uncertainty spatially, kriging wins.
